Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-01 article-models-tt.tex
second proj of x.u is u
whoops, I proved the last two things in the wrong place. Hehe
worked on extension on terms, initiated two new sections

\subsection{Extension on terms}\label{extension-on-terms}
When $\jterm{\Gamma}{A}{x}$ and $\jterm{\Gamma}{\subst{x}{P}}{u}$ are terms,
we obtain the term $\jterm{\Gamma}{\ctxext{A}{P}}{\subst{u}{{x}{\idfunc[\ctxext{A}{P}]}}}$. 
We define 
\begin{equation*}
\ctxext{x}{u}\defeq\subst{u}{{x}{\idfunc[\ctxext{A}{P}]}}.
\end{equation*}
Thus
$\ctxext{x}{u}$ is the pairing of $x$ and $u$. Note that because
$\ctxwk{P}{{A}{\ctxext{A}{P}}}\jdeq\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}$, the
pairing function is just $\idfunc[\ctxext{A}{P}]$.

When we substitute by an extended term we get an equal result as when we
substitute two consecutive times, like the way currying works. Note however,
that we can now \emph{prove} this property using the inference rules we have
already established. It is not a matter of convention, the way it usually is
in type theory by agreeing that $f(x)(y)$ is written as $f(x,y)$.

@@

There is also a decomposition of terms $\jterm{\Gamma}{\ctxext{A}{P}}{w}$ into
components: the projections. We define
\begin{align*}
\cprojfstf{A}{P} & \defeq \ctxwk{P}{\idfunc[A]}\\
\cprojsndf{A}{P} & \defeq \idfunc[P]
\end{align*}

@@

\begin{proof}
Let $\jfam{\ctxext{\Gamma}{A}}{P}$ and $\jfam{\ctxext{\Gamma}{A}}{Q}$ be
families. Then we have
\begin{align*}
\subst{\cprojfstf{A}{P}}{\ctxwk{\ctxext{A}{P}}{Q}} 
& \jdeq \subst{\ctxwk{P}{\idfunc[A]}}{\ctxwk{\ctxext{A}{P}}{Q}} \tag{by definition}\\
& \jdeq \subst{\ctxwk{P}{\idfunc[A]}}{\ctxwk{P}{{A}{Q}}} \tag{by \autoref{comp-ww-f}}\\
& \jdeq \ctxwk{P}{\subst{\idfunc[A]}{\ctxwk{A}{Q}}} \tag{by \autoref{comp-ws-f}}\\
& \jdeq \ctxwk{P}{Q} \tag{by \autoref{idfunc-wk-defn}}
\end{align*}
To prove the second judgmental equality, let $\jterm{\Gamma}{A}{x}$ and
$\jterm{\Gamma}{\subst{x}{P}}{u}$. Then we have
\begin{align*}
\cprojfst{A}{P}{\ctxext{x}{u}}
& \jdeq \subst{\ctxext{x}{u}}{\ctxwk{P}{\idfunc[A]}}\\
& \jdeq \subst{u}{{x}{\ctxwk{P}{\idfunc[A]}}} \\
& \jdeq \subst{u}{\ctxwk{\subst{x}{P}}{\subst{x}{\idfunc[A]}}}\\
& \jdeq \subst{x}{\idfunc[A]}\\
& \jdeq x.
\end{align*}
\end{proof}

@@

\begin{proof}
To prove the first judgmental equality, note that
\begin{align*}
w & \jdeq \subst{w}{\idfunc[\ctxext{A}{P}]} \tag{by \autoref{idfunc-subst-defn}}\\
& \jdeq \subst{w}{\subst{\idfunc[P]}{{\ctxwk{P}{\idfunc[A]}}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}}
  \tag{by \autoref{idfunc-ext-comp}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{\subst{w}{{\ctxwk{P}{\idfunc[A]}}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}}\tag{by \autoref{comp-ss-t}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{{\subst{w}{\ctxwk{P}{\idfunc[A]}}}{\subst{w}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}}\tag{by \autoref{comp-ss-t}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{{\subst{w}{\ctxwk{P}{\idfunc[A]}}}{\subst{w}{\ctxwk{\ctxext{A}{P}}{\idfunc[\ctxext{A}{P}]}}}}\tag{by \autoref{idfunc-wk-comp}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{{\subst{w}{\ctxwk{P}{\idfunc[A]}}}{\idfunc[\ctxext{A}{P}]}}\tag{by \autoref{defn-ws-4}}\\
& \jdeq \ctxext{\cprojfst{A}{P}{w}}{\cprojsnd{A}{P}{w}}\tag{by definition}
\end{align*}
\end{proof}

@@

@@

>>>>

\subsection{Extension on terms}\label{extension-on-terms}
In this subsection we consider the notion of extension on terms, which has now
become definable inside our theory. Moreover, every compatibility rule one may
dream of is provable as well, using the compatibility rules we have introduced
earlier. \emph{This subsection contains no new assumptions.}

\begin{defn}
When $\jterm{\Gamma}{A}{x}$ and $\jterm{\Gamma}{\subst{x}{P}}{u}$ are terms,
we define 
\begin{equation*}
\jtermdefn{\Gamma}{\ctxext{A}{P}}{\ctxext{x}{u}}{\subst{u}{{x}{\idfunc[\ctxext{A}{P}]}}}.
\end{equation*} 
\end{defn}

Thus, the term $\ctxext{x}{u}$ is the pairing of $x$ and $u$. Note that because
we have the judgmental equality 
$\ctxwk{P}{{A}{\ctxext{A}{P}}}\jdeq\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}$ in the
context $\ctxext{{\Gamma}{A}}{P}$, the
pairing function could just be defined as $\idfunc[\ctxext{A}{P}]$.

When we substitute by an extended term we get an equal result as when we
substitute two consecutive times, like the way currying works.

@@

There is also a decomposition of terms $\jterm{\Gamma}{\ctxext{A}{P}}{w}$ into
components: the projections.

\begin{defn}
Let $\jfam{\Gamma}{A}$ and $\jfam{\ctxext{\Gamma}{A}}{P}$ be families. We define
\begin{align*}
\jtermdefn*{\ctxext{\Gamma}{{A}{P}}}{\ctxwk{P}{{A}{A}}}{\cprojfstf{A}{P}}{\ctxwk{P}{\idfunc[A]}}\\
\jtermdefn*{\ctxext{\Gamma}{{A}{P}}}{\ctxwk{P}{P}}{\cprojsndf{A}{P}}{\idfunc[P]}
\end{align*}
\end{defn}

@@

\begin{proof}
Let $\jfam{\ctxext{\Gamma}{A}}{P}$ and $\jfam{\ctxext{\Gamma}{A}}{Q}$ be
families. Then we have
\begin{align*}
\subst{\cprojfstf{A}{P}}{\ctxwk{\ctxext{A}{P}}{Q}} 
& \jdeq \subst{\ctxwk{P}{\idfunc[A]}}{\ctxwk{\ctxext{A}{P}}{Q}} \tag{by definition}\\
& \jdeq \subst{\ctxwk{P}{\idfunc[A]}}{\ctxwk{P}{{A}{Q}}} \tag{by \autoref{comp-ww-f}}\\
& \jdeq \ctxwk{P}{\subst{\idfunc[A]}{\ctxwk{A}{Q}}} \tag{by \autoref{comp-ws-f}}\\
& \jdeq \ctxwk{P}{Q} \tag{by \autoref{idfunc-wk-defn}}
\end{align*}
\end{proof}

@@

\begin{proof}
To prove the first judgmental equality, note that
\begin{align*}
w & \jdeq \subst{w}{\idfunc[\ctxext{A}{P}]} \tag{by \autoref{idfunc-subst-defn}}\\
& \jdeq \subst{w}{\subst{\idfunc[P]}{{\ctxwk{P}{\idfunc[A]}}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}}
  \tag{by \autoref{idfunc-ext-comp}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{\subst{w}{{\ctxwk{P}{\idfunc[A]}}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}}\tag{by \autoref{comp-ss-t}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{{\subst{w}{\ctxwk{P}{\idfunc[A]}}}{\subst{w}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}}\tag{by \autoref{comp-ss-t}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{{\subst{w}{\ctxwk{P}{\idfunc[A]}}}{\subst{w}{\ctxwk{\ctxext{A}{P}}{\idfunc[\ctxext{A}{P}]}}}}\tag{by \autoref{idfunc-wk-comp}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{{\subst{w}{\ctxwk{P}{\idfunc[A]}}}{\idfunc[\ctxext{A}{P}]}}\tag{by \autoref{defn-ws-4}}\\
& \jdeq \ctxext{\cprojfst{A}{P}{w}}{\cprojsnd{A}{P}{w}}\tag{by definition}
\end{align*}
To prove the second judgmental equality, let $\jterm{\Gamma}{A}{x}$ and
$\jterm{\Gamma}{\subst{x}{P}}{u}$. Then we have
\begin{align*}
\cprojfst{A}{P}{\ctxext{x}{u}}
& \jdeq \subst{\ctxext{x}{u}}{\ctxwk{P}{\idfunc[A]}}\\
& \jdeq \subst{u}{{x}{\ctxwk{P}{\idfunc[A]}}} \\
& \jdeq \subst{u}{\ctxwk{\subst{x}{P}}{\subst{x}{\idfunc[A]}}}\\
& \jdeq \subst{x}{\idfunc[A]}\\
& \jdeq x.
\end{align*}
To prove the third judgmental equality, note that
\begin{align*}
\cprojsnd{A}{P}{\ctxext{x}{u}}
& \jdeq \subst{\ctxext{x}{u}}{\idfunc[P]}\\
& \jdeq \subst{u}{{x}{\idfunc[P]}}\\
& \jdeq \subst{u}{\idfunc[\subst{x}{P}]}\\
& \jdeq u.
\end{align*}
\end{proof}

@@

\begin{proof}
Consider terms $\jterm{\Gamma}{A}{x}$, $\jterm{\Gamma}{\subst{x}{P}}{u}$ and
$\jterm{\Gamma}{\subst{u}{{x}{Q}}}{v}$. Then we have
\begin{align*}
\ctxext{x}{{u}{v}} & \jdeq \subst{\ctxext{u}{v}}{{x}{\idfunc[\ctxext{A}{{P}{Q}}]}}\\
& \jdeq \subst{v}{{u}{{x}{\idfunc[\ctxext{A}{{P}{Q}}]}}}\\
& \jdeq \subst{v}{{u}{{x}{\idfunc[\ctxext{{A}{P}}{Q}]}}}\\
& \jdeq \subst{v}{{\ctxext{x}{u}}{\idfunc[\ctxext{{A}{P}}{Q}]}}\\
& \jdeq \ctxext{{x}{u}}{v}.
\end{align*}
Now consider a term $\jterm{\Gamma}{\ctxext{A}{{P}{Q}}}{w}$. Then we have
\begin{align*}
w 
& \jdeq \ctxext{\cprojfst{A}{\ctxext{P}{Q}}{w}}{\cprojsnd{A}{\ctxext{P}{Q}}{w}}\\
& \jdeq \ctxext{\cprojfst{A}{\ctxext{P}{Q}}{w}}{{\cprojfst{P}{Q}{\cprojsnd{A}{\ctxext{P}{Q}}{w}}}{\cprojsnd{P}{Q}{\cprojsnd{A}{\ctxext{P}{Q}}{w}}}}\\
& \jdeq \ctxext{{\cprojfst{A}{\ctxext{P}{Q}}{w}}{\cprojfst{P}{Q}{\cprojsnd{A}{\ctxext{P}{Q}}{w}}}}{\cprojsnd{P}{Q}{\cprojsnd{A}{\ctxext{P}{Q}}{w}}}
\end{align*}
Thus we see that 
\begin{align*}
\cprojfst{\ctxext{A}{P}}{Q}{w} & \jdeq \ctxext{\cprojfst{A}{\ctxext{P}{Q}}{w}}{\cprojfst{P}{Q}{\cprojsnd{A}{\ctxext{P}{Q}}{w}}}\\ 
\cprojsnd{\ctxext{A}{P}}{Q}{w} & \jdeq \cprojsnd{P}{Q}{\cprojsnd{A}{\ctxext{P}{Q}}{w}},
\end{align*}
proving the fourth judgmental equality, and therefore also that
\begin{align*}
\cprojfst{A}{P}{\cprojfst{\ctxext{A}{P}}{Q}{w}} & \jdeq \cprojfst{A}{\ctxext{P}{Q}}{w}\\
\cprojsnd{A}{P}{\cprojfst{\ctxext{A}{P}}{Q}{w}} & \jdeq \cprojfst{P}{Q}{\cprojsnd{A}{\ctxext{P}{Q}}{w}},
\end{align*}
proving the second and the third judgmental equalities.
\end{proof}

@@

\section{Pretty type theory}
In this section we will do three things. First we explain basic type theory with
variable names, which we simply call \emph{Pretty Type Theory}. 
Then we will show how every formula in basic type theory without
variable names (shall we call it \emph{Structural Type Theory}?) 
can be interpreted in pretty type theory. Finally, we will show how every formula
in pretty type theory can be interpreted in structural type theory.

\section{Chalmers type theory}
In this section we will present the type theory that Coquand and Dybjer are using.
It is a weak type theory (I think), with not so many operations and judgmental equalities.
We will show how every formula of Chalmers type theory can be interpreted in
structural type theory.
